A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2016-11-12 linear.tex
minor structural change and edits

\subsection{Numerical examples}

@@

\subsection{Comparison to the convective overstability}
%These slowly-growing modes have $|\omega|\simeq \kappa$, so they correspond to
%overstable epicycles. 
The streaming
instability bears some similarity  to the `convective 
overstability' in dust-free, non-isothermal disks 
\citep{klahr14,lyra14,latter16}. In that case, the combination of an 
unstable radial buoyancy frequency ($N_r^2 < 0 $) and
finite thermal relaxation time leads to growing epicyclics. %with
%growth rate $\propto  -N_r^2$

Indeed, the effective radial buoyancy frequency in our dusty disk 
with negligble background density gradient is unstable, 
\begin{align}
  N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
  -\frac{\rho}{P}F_r^2<0.  
\end{align}
Convective overstability cannot exist in our locally isothermal disks,
which have zero cooling times. On the other hand, dust-gas drag 
is equivalent to heating/cooling in our framework.

One might then think of the streaming instability as a dusty relative
to the convective  overstability. Explicit solutions in the small
$\tstop$ limit show that streaming instability growth rates are
also $\propto |N_r^2|$  \citep{jacquet11}, as is for the convective
overstability \citep{lyra14,latter16}. Furthermore, both instabilities require
finite effective cooling times, and both require $k_z\neq0$. 
%{\bf note: si osc freq not too small compared to kappa}

However, the convective overstability is most unstable at $k_x=0$
\citep{lyra14}  whereas the streaming instability growth rate is maximized at finite $k_x$
\citep{youdin05a}. Nevertheless, they share the same underlying
physical mechanism, i.e. pressure-density phase lags.

>>>>

%\subsection{Numerical examples}

@@

### 2016-11-12 model.tex
minor structural change and edits

In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real integral $\mathcal{I}^2>0$ and coefficients $A,B,C,D$ can be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To see this, we first compute the work done
assuming periodic oscillations. We then show that if the work done is positive, then the
oscillation amplitude in fact grows: the work done leads the
system to `overshoot' past the amplitude of the preceeding oscillation
cycle.

%its equilibrium. {\bf should be overshoot the max displacement}

The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{D\upsilon}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep{cox67}, 
where $T_p$ is the oscillation period, $\upsilon=1/\rho$ is the specific
volume of the mixture, and the spatial integral is taken over the
total mass $M$ of the mixture and $dm$ is the mass element. 
Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if gas pressure lags behind density, then %positive work done 
$\mathcal{W}>0$ because it leads to a clockwise path in the
`P-V' plane. From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
%{\bf should be: expand beyond the max
%  vol of the previous cycle}
%corresponds to positive work done. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
% {\bf
  %should be contract beyond the min vol of prev cycle}
The 
overall positive work done leads to growth in the oscillation amplitude.

@@

%Applying the above argument to isothermal dusty gas, 
For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of overstable modes caused by
    dust-gas drag. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown,
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the P-V plane, implying positive work done by the fluid and
    hence would lead to growing oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

>>>>

In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
%&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
  \sigma^2\mathcal{I}^2
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align} 
%The real integral $\mathcal{I}^2>0$ 
where
\begin{align}
  \mathcal{I}^2 \equiv \int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV
\end{align}
is the meridional kinetic energy, 
and coefficients $A,B,C,D$ can be
read off Eq. \ref{integral_ex}. %Note that $B=C$ from dynamical equilibrium
%(Eq. \ref{vshear}). 
We now consider various limits of Eq. \ref{int_rel}.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done during a periodic oscillation, then show that if the work
done is positive, then the growth rate is positive, i.e. the
oscillation amplitude would actually grow.

Consider oscillations in the dusty gas with period $T_p$. 
The average rate of work done is 
\begin{align}
  \mathcal{W} = \frac{1}{T_p}\int^{t+T_p}_{t}dt^\prime\int_M P
  \frac{D\upsilon}{Dt^\prime} dm \label{work_def} 
\end{align}
\citep{cox67}, 
where $\upsilon=1/\rho$ is the specific volume of the mixture, $D/Dt$
is the Lagragian derivative, and the 
spatial integral is taken over the total mass $M$ of the mixture and
$dm$ is a mass element.

Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

@@

@@

%Applying the above argument to isothermal dusty gas,

For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components.

@@

### 2016-11-12 paper.tex
adding kz=0 special case in response to inutsuka-san's query
general clean up
minor structural change and edits

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\rich}{\mathrm{Ri}}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{f_\mathrm{d}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}}
\newcommand{\seff}{S_\mathrm{eff}}
\newcommand{\tcool}{t_\mathrm{cool}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks:
  axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is analogous to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
However, sufficiently sharp radial gradients in the dust-to-gas ratio
can destabilize the disk.  
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also duscuss thermodynamical interpretations of 
dust-gas drag instabilities, and show that the streaming
instability in protoplanetary disks occur because drag causes 
a phase lag between the gas pressure  
and dust density, much like pulsational instabilities 
in stars.  
\end{abstract}

@@

This paper is organized as follows. In \S\ref{setup} we list the
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} which are used in
explicit calculations. In \S\ref{limits} we discuss
general stability properties of dusty disks. We then move onto
explicit stablity analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} study dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

>>>>

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\rich}{\mathrm{Ri}}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{f_\mathrm{d}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\taus}{\tau_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}}
\newcommand{\seff}{S_\mathrm{eff}}
\newcommand{\tcool}{t_\mathrm{cool}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to show that 
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, and can be understood as a result of dust-induced
  buoyancy forces. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
equivalence in finding dusty analogs of other hydrodynamic
instabilities, as well as a way to simulate dusty protoplanetary disks
using purely hydrodynamic codes with minimal effort.

\end{abstract}

@@

This paper is organized as follows. We first explain in  
\S\ref{grow_osc} how a pressure-density phase lag in fluids can lead to
instability. This motives us to find an analogy between dusty gas and
pure hydrodynamics, as it would allow us to understand dust dynamics
within the framework of standard hydrodynamics.

We begin our formal discussion in in \S\ref{setup} by listing the 
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} which are used in
explicit calculations. In \S\ref{limits} we discuss
general stability properties of dusty disks. We then move onto
explicit stablity analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} study dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever the
pressure and density evolution of the fluid are not in phase. If 
the average work done is positive, then oscillation amplitudes 
grow: the work done allows the system to `overshoot' beyond
the amplitude of preceeding oscillations.

%To see this, we first compute the work done
%assuming periodic oscillations. We then show that if the work done is positive, then the
%oscillation amplitude in fact grows: the work done leads the
%system to `overshoot' past the amplitude of the preceeding oscillation
%cycle. 
%its equilibrium. {\bf should be overshoot the max displacement}

Fig. \ref{pdv_cartoon} gives a graphical demonstration that 
if pressure lags behind density, then positive is work done 
because it leads to a clockwise path in the 
`P-V' plane. The annotations consider the particular case where the
phase lag arises because the fluid has two components (gas and dust,
see below) but the following description is general.

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing on the way
down.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of growing oscillations in a
    fluid. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown, 
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the P-V plane, implying positive work done by the fluid and
    hence would lead to growing oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature is time-independent but varies with radius
\citep{chiang97}. Then the pressure of a fluid parcel varies with 
position as well. There is no reason to expect $P$ and $\rho$ to
be in phase. In fact, we will show that this is responsible for 
the vertical shear instability \citep{lin15}.

%temperature of a fluid parcel changes with position

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not neccessarily in phase. If this causes $P$ to lag
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

In order to apply this thermodynamic interpretation of dust-gas drag
instabilities, we need to develop an analogy between dusty-gas
and pure gas dynamics. We show this is possible in the limit of
strong drag and a fixed gas equation of state.

@@

%{\bf acknowledgements. s. inutsuka for asking about kz=0 limit}

### 2016-11-12 summary.tex
minor structural change and edits

>>>>

\subsubsection{Convective overstability}
The `convective overstability' (ConO) operates in dust-free, non-adiabatic disks 
when there is an unstable radial buoyancy frequency ($N_r^2 < 0 $)
combined with finite thermal relaxation time, $t_\mathrm{cool}\sim
\OmK^{-1}$. This lead to growing epicycles
\citep{klahr14,lyra14,latter16}.  
%with 
%growth rate $\propto  -N_r^2$ 
% Indeed, the effective radial buoyancy frequency in our dusty disk 
% with negligble background density gradient is unstable, 
% \begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

The conditions required for ConO might be realized in a locally
isothermal dusty disk. Since $\p_rP < 0$ in typical disk models, 
if the dust-induced radial entropy gradient is negative, then
$N_r^2  =  F_r \p_r\seff < 0$. In 
fact, typical disk models used for the streaming
instability have negligible background radial density gradient, then  
$N_r^2= -\rho F_r^2/P$.

Locally isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture. It would be
interesting to explore whether or not dust-gas drag can provide the
required `cooling' to enable a dusty version of the convective
overstability.

%Furthermore, 
%the effective radial buoyancy frequency in our dusty disk  
%with negligble background density gradient is unstable, 
%\begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

% One might then think of the streaming instability as a dusty relative
% to the convective  overstability. Explicit solutions in the small
% $\tstop$ limit show that streaming instability growth rates are
% also $\propto |N_r^2|$  \citep{jacquet11}, as is for the convective
% overstability \citep{lyra14,latter16}. Furthermore, both instabilities require
% finite effective cooling times, and both require $k_z\neq0$. 
% %{\bf note: si osc freq not too small compared to kappa}

% However, the convective overstability is most unstable at $k_x=0$
% \citep{lyra14}  whereas the streaming instability growth rate is maximized at finite $k_x$
% \citep{youdin05a}.

\subsubsection{Zombie vortices}

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear (or finite-amplitude) instability in gas disks
\citep{marcus15}. It operates by exciting `critical layers' which role
up into vortices, exciting further critical layers. The result is an
array of vortices. These critical
layers are locations where the intrinsic wave frequency of a
disturbance matches the local vertical buoyancy frequency
\citep{marcus13}.

Since buoyancy is critical for the ZVI, it requires an almost
purely adiabatic gas ($\tcool\to\infty$). This limits its
applicability to the inner most parts of typical protoplanetary disks
\citep{lesur16}.

However, we have show that dust-loading an isothermal disk also
induces an effective buoyancy. This effect is maximized when dust is
perfectly coupled to the gas. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'.
